---
title: "Installation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

This vignette describes a reproducible installation workflow for `STID`. Because `STID` integrates spatial transcriptomics, niche detection, cell-cell communication, regulatory-network inference, and host-pathogen analysis, the package depends on a relatively broad R, Bioconductor, GitHub, and system-software ecosystem.

This vignette covers:

-   preparing a clean R library for `STID`;
-   configuring CRAN and Bioconductor mirrors;
-   installing core R, Bioconductor, and GitHub dependencies;
-   installing `STID` from GitHub;
-   checking whether required and suggested packages are available;
-   troubleshooting common installation failures caused by package versions, compilation tools, network access, or system libraries.

> **Note:** Before running the commands below, update the CRAN mirror, Bioconductor mirror, library path, R version, compiler settings, and optional dependency list according to the operating system and network environment used for the analysis.

## Installation strategy

A controlled installation environment is recommended. The general workflow is:

1.  confirm that R and system build tools are available;
2.  configure a dedicated library path for `STID`;
3.  configure CRAN and Bioconductor mirrors;
4.  install package-management utilities;
5.  install heavy or failure-prone dependencies first;
6.  install `STID` from GitHub;
7.  verify core and optional dependencies.

> **Note:** If an older package version is already loaded from the default library, simply calling `.libPaths("./library_STID/")` may not fully solve version conflicts. In that case, restart R and update or remove the conflicting package from the default library before reinstalling `STID`.


## Configure the R installation environment

### Configure mirrors

Use a CRAN mirror and Bioconductor mirror that are stable in the current network environment.

```{r configure-mirrors, eval = FALSE}
# Recommended for users in China
options(repos = c(CRAN = "https://mirrors.tuna.tsinghua.edu.cn/CRAN/"))
options(BioC_mirror = "https://mirrors.tuna.tsinghua.edu.cn/bioconductor")

# Recommended for users outside China
options(repos = c(CRAN = "https://cloud.r-project.org"))
options(BioC_mirror = "https://bioconductor.org")
```

### Configure an isolated library path

A dedicated library path helps avoid conflicts with an existing R environment and makes the installation easier to reproduce.

```{r configure-library-path, eval = FALSE}
dir.create("./library_STID/", showWarnings = FALSE, recursive = TRUE)
.libPaths(normalizePath("./library_STID/"))
.libPaths()
```

> **Note:** If older dependency versions already exist in the default library (for example, `Matrix`, `rlang`, or `SeuratObject`), version conflicts may still occur during installation. It is recommended to restart R and prioritize upgrading the conflicting packages in the default library, or reinstall everything in a fresh R environment.

### Configure download and compilation options

```{r configure-installation-options, eval = FALSE}
# Increase download timeout for large packages or unstable networks
options(timeout = 300)

# Use parallel compilation when source installation is required
n_cores <- max(1, parallel::detectCores() - 2)
Sys.setenv(MAKEFLAGS = paste0("-j", n_cores))

# Reduce debug information and use optimized compilation flags
Sys.setenv(
  PKG_CFLAGS = "-O2 -g0 -DNDEBUG",
  PKG_CXXFLAGS = "-O2 -g0 -DNDEBUG"
)
```

## Install package-management utilities

Install `remotes`, `devtools`, and `BiocManager` before installing `STID` and its dependencies.

```{r install-package-managers, eval = FALSE}
install_if_missing <- function(pkgs) {
  missing_pkgs <- pkgs[!vapply(pkgs, requireNamespace, logical(1), quietly = TRUE)]
  if (length(missing_pkgs) > 0) {
    install.packages(missing_pkgs)
  }
  invisible(missing_pkgs)
}

install_if_missing(c("remotes", "devtools", "BiocManager"))
```

## Install core dependencies

The following packages are commonly involved in version or compilation failures. Installing them explicitly before `STID` can make the GitHub installation more stable.

> **Note:** For recent R versions, binary packages are usually available for many dependencies and can substantially reduce installation time. For older R versions, source installation or version pinning may be required.

```{r install-core-cran-dependencies, eval = FALSE}
install.packages(c(
  "rlang",
  "openssl",
  "curl",
  "systemfonts"
))

install.packages(c(
  "Matrix",
  "RcppArmadillo",
  "uwot",
  "units",
  "ggforce"
), type = "binary")
```

If the binary packages are unavailable or incompatible with the current R version, install a compatible source version manually.

```{r install-version-pinned-dependencies, eval = FALSE}
remotes::install_version("rlang", version = "1.1.7")
remotes::install_version("Matrix", version = "1.6.5")
remotes::install_version("SeuratObject", version = "5.0.2")
remotes::install_version("Seurat", version = "5.1.0")
```

### Bioconductor dependencies

```{r install-bioconductor-dependencies, eval = FALSE}
BiocManager::install(c(
  "AnnotationDbi",
  "AUCell",
  "Biostrings",
  "clusterProfiler",
  "DOSE",
  "GO.db",
  "GOSemSim",
  "SingleR",
  "STRINGdb",
  "SummarizedExperiment",
  "UCell"
), ask = FALSE, update = FALSE)
```

### GitHub dependencies

Some downstream analyses use packages that are usually installed from GitHub. Install these packages only if the corresponding analysis modules are needed.

```{r install-github-dependencies, eval = FALSE}
devtools::install_github("jinworks/CellChat", upgrade = "never")
devtools::install_github("saezlab/mistyR", upgrade = "never")
devtools::install_github("aeyslab/nichenetr", upgrade = "never")
devtools::install_github("mhahsler/rBLAST", upgrade = "never")
devtools::install_github("mojaveazure/seurat-disk#198", upgrade = "never")
```

> **Note:** When GitHub installation fails with `Error in download.file`, the most common causes are network interruption, rate limits, or proxy configuration. Retry the command, configure a proxy if needed, or install the package from a downloaded source archive.

## Install STID

Install `STID` directly from GitHub. When prompted to update dependencies, selecting `3: None` is often safer in an already configured environment because it avoids unexpectedly overwriting packages such as `Matrix`, `SeuratObject`, or `Seurat`.

```{r install-stid, eval = FALSE}
remotes::install_github(
  "YulongQin/STID",
  dependencies = TRUE,
  upgrade = "never"
)
```

The full installation may take 30--60 minutes or longer depending on the operating system, R version, network speed, and whether packages are installed from source.

```{r load-stid, eval = FALSE}
library(STID)
packageVersion("STID")
```

## Check dependencies

The `Imports` packages are required for core `STID` functions. Packages in `Suggests` are not always installed automatically and should be installed manually before running the corresponding downstream analysis.

```{r define-stid-dependencies, eval = FALSE}
pkg_imports <- c(
  "anndata", "AnnotationDbi", "assertthat", "AUCell", "Biostrings",
  "clusterProfiler", "concaveman", "data.table", "dbscan", "distances",
  "doParallel", "DOSE", "dplyr", "DT", "FNN", "foreach", "furrr",
  "future", "ggdensity", "ggforce", "ggh4x", "ggnewscale", "ggplot2",
  "ggraph", "ggrepel", "ggsignif", "ggvenn", "GO.db", "GOSemSim",
  "igraph", "imager", "magrittr", "Matrix", "methods", "paletteer",
  "patchwork", "proxy", "psych", "purrr", "rBLAST", "reticulate",
  "rjson", "rlang", "scales", "Seurat", "SeuratDisk", "SeuratObject",
  "shiny", "shinycssloaders", "sp", "STRINGdb", "stringr",
  "SummarizedExperiment", "tibble", "tidyr", "UCell", "viridis", "zip", "Mfuzz"
)

pkg_suggests <- c(
  "CellChat", "knitr", "mistyR", "nichenetr", "org.Hs.eg.db",
  "org.Mm.eg.db", "rmarkdown", "SingleR", "testthat"
)

pkg_min_versions <- c(
  ggplot2 = "4.0.2",
  rlang = "1.1.7",
  Matrix = "1.6.5",
  SeuratObject = "5.0.2",
  Seurat = "5.1.0"
)
```

```{r check-stid-dependencies, eval = FALSE}
check_stid_packages <- function(pkg_list, min_versions = NULL) {
  installed_pkgs <- rownames(installed.packages())
  installed_ok <- intersect(pkg_list, installed_pkgs)
  missing_pkgs <- setdiff(pkg_list, installed_pkgs)

  version_wrong <- character()
  if (!is.null(min_versions)) {
    for (pkg in names(min_versions)) {
      if (pkg %in% installed_ok) {
        current_version <- as.character(packageVersion(pkg))
        required_version <- min_versions[[pkg]]
        if (utils::compareVersion(current_version, required_version) < 0) {
          version_wrong <- c(
            version_wrong,
            sprintf(
              "%s (requires >= %s, current = %s)",
              pkg, required_version, current_version
            )
          )
        }
      }
    }
  }

  cat("==================== Package Installation Check ====================\n")
  cat("Installed packages: ", length(installed_ok), "\n", sep = "")
  cat("Missing packages: ", length(missing_pkgs), "\n", sep = "")

  if (length(missing_pkgs) > 0) {
    cat("\nMissing list:\n")
    cat(paste0("  - ", missing_pkgs, "\n"), sep = "")
  } else {
    cat("\nAll packages in this list are installed.\n")
  }

  if (length(version_wrong) > 0) {
    cat("\nVersion mismatches:\n")
    cat(paste0("  - ", version_wrong, "\n"), sep = "")
  }
  cat("====================================================================\n")

  invisible(list(
    installed = installed_ok,
    missing = missing_pkgs,
    version_mismatch = version_wrong
  ))
}

imports_check <- check_stid_packages(pkg_imports, pkg_min_versions)
suggests_check <- check_stid_packages(pkg_suggests)
```

### Install missing dependencies from the check result

Use the dependency check results to install missing packages in smaller batches. This makes it easier to identify which package fails.

```{r install-missing-cran-packages, eval = FALSE}
missing_imports <- imports_check$missing
missing_suggests <- suggests_check$missing

cran_candidates <- setdiff(
  unique(c(missing_imports, missing_suggests)),
  c(
    "AnnotationDbi", "AUCell", "Biostrings", "clusterProfiler", "DOSE",
    "GO.db", "GOSemSim", "SingleR", "STRINGdb", "SummarizedExperiment",
    "UCell", "CellChat", "mistyR", "nichenetr", "rBLAST", "SeuratDisk",
    "org.Hs.eg.db", "org.Mm.eg.db"
  )
)

if (length(cran_candidates) > 0) {
  install.packages(cran_candidates)
}
```

```{r install-missing-bioconductor-packages, eval = FALSE}
bioc_candidates <- intersect(
  unique(c(missing_imports, missing_suggests)),
  c(
    "AnnotationDbi", "AUCell", "Biostrings", "clusterProfiler", "DOSE",
    "GO.db", "GOSemSim", "SingleR", "STRINGdb", "SummarizedExperiment",
    "UCell", "org.Hs.eg.db", "org.Mm.eg.db"
  )
)

if (length(bioc_candidates) > 0) {
  BiocManager::install(bioc_candidates, ask = FALSE, update = FALSE)
}
```

## Optional analysis modules

Some `STID` functions require additional resources beyond the core R package installation.

### Cell-cell communication analysis

`CalSampCellComm()` requires `CellChat` and ligand-receptor resources. Install and test `CellChat` before running communication analysis.

```{r check-cellchat, eval = FALSE}
if (!requireNamespace("CellChat", quietly = TRUE)) {
  devtools::install_github("jinworks/CellChat", upgrade = "never")
}
```

### Spatial colocalization analysis

`CalSampCoLoc()` can use `mistyR` or Python-based tools depending on the selected method. If the workflow uses Python packages, configure `reticulate` to point to the intended Python environment.

```{r configure-reticulate, eval = FALSE}
library(reticulate)

# Replace this path with the Python or conda environment used for spatial analysis
# use_python("/path/to/python", required = TRUE)
# use_condaenv("stid", required = TRUE)

py_config()
```

### Gene regulatory network analysis

`CalSampGRN()` requires NicheNet reference files and organism-specific resources. Download and store these files locally before running the analysis.

```{r check-nichenet, eval = FALSE}
if (!requireNamespace("nichenetr", quietly = TRUE)) {
  devtools::install_github("aeyslab/nichenetr", upgrade = "never")
}
```

### Host-pathogen protein association analysis

`CalSampPPI()` requires local FASTA files, symbol-to-protein mapping files, BLAST, and STRING resources.

```{r check-blast, eval = FALSE}
Sys.which("blastp")
Sys.which("makeblastdb")

if (!requireNamespace("rBLAST", quietly = TRUE)) {
  devtools::install_github("mhahsler/rBLAST", upgrade = "never")
}
```

## Troubleshooting

### Package version conflicts

Version conflicts commonly involve `Matrix`, `rlang`, `SeuratObject`, `Seurat`, or `ggplot2`.

```{r inspect-package-versions, eval = FALSE}
packages_to_check <- c("Matrix", "rlang", "SeuratObject", "Seurat", "ggplot2")

vapply(packages_to_check, function(pkg) {
  if (requireNamespace(pkg, quietly = TRUE)) {
    as.character(packageVersion(pkg))
  } else {
    NA_character_
  }
}, character(1))
```

> **Note:** Restart R after updating core packages. A package namespace that is already loaded in the current R session may continue to use the old version until the session is restarted.

### Network or GitHub download failures

```{r network-troubleshooting, eval = FALSE}
options(timeout = 600)

# Test whether GitHub is reachable from R
utils::download.file(
  "https://raw.githubusercontent.com/YulongQin/STID/main/DESCRIPTION",
  tempfile(),
  quiet = FALSE
)
```

If the download fails, retry later, switch networks, configure a proxy, or download the source archive manually and install from the local file.

```{r install-from-local-source, eval = FALSE}
# Example: install from a local source archive
# remotes::install_local("/path/to/STID-main.zip", dependencies = TRUE, upgrade = "never")
```

### Compilation failures

Compilation failures usually indicate missing build tools or system libraries. Install the corresponding system dependency, restart R, and reinstall the failed package.

```{r reinstall-failed-package, eval = FALSE}
# Replace units with the package that failed in your environment
remove.packages("units")
install.packages("units", type = "source")
```

### Bioconductor version mismatch

Bioconductor packages must match the active R version. Confirm the Bioconductor version before installing Bioconductor dependencies.

```{r check-bioc-version, eval = FALSE}
BiocManager::version()
BiocManager::valid()
```

## Notes

-   Use a clean library path and restart R before installation to avoid stale namespaces.
-   Use R 4.5 or higher when possible to reduce source-compilation and dependency-resolution problems.
-   Install heavy or failure-prone dependencies before installing `STID` from GitHub.
-   Suggested packages are not always installed automatically; install them manually before running optional downstream analyses.
-   For reproducible projects, record the R version, package versions, system libraries, and installation date.

## Session information

```{r session-info}
sessionInfo()
```
