---
title: "Datasets and gene sets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Datasets and gene sets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  fig.retina = 2
)

set.seed(123)
```

## Overview

This vignette briefly summarizes the public spatial transcriptomic datasets and curated gene resources used by `STID`.

```{r packages, message = FALSE, warning = FALSE, eval = FALSE}
library(tidyverse)
library(Seurat)
library(STID)
```

## Public datasets

Table S1 includes eight public mouse spatial transcriptomic datasets covering viral, bacterial, and parasitic infection models.

| Pathogen type | Pathogen | Abbrev. | Disease | Tissue | Reference | Technology | GEO | n |
|---|---|---|---|---|---|---|---|---|
| Virus | Japanese encephalitis virus | JEV | Japanese encephalitis | Brain | Ou et al. (2026) | Stereo-seq | N/A | 12 |
| Virus | Venezuelan equine encephalitis virus | VEEV | Venezuelan equine encephalitis | Brain | Rangel et al. (2024) | 10x Visium | GSE275201 | 2 |
| Virus | Reovirus type 1 Lang | Reo-T1L | Viral myocarditis | Heart | Mantri et al. (2022) | 10x Visium | GSE189636 | 4 |
| Bacterium | Mycobacterium tuberculosis | MTB | Tuberculosis | Lung | Zhao et al. (2025) | Stereo-seq V2 | N/A | 8 |
| Bacterium | Klebsiella pneumoniae | Kp | Klebsiella pneumoniae pneumonia | Lung | Xu et al. (2022) | 10x Visium | GSE190225 | 4 |
| Parasite | Echinococcus multilocularis | Em | Alveolar echinococcosis | Liver | Ou et al. (2025) | Stereo-seq | N/A | 12 |
| Parasite | Plasmodium berghei ANKA | N/A | Malaria | Liver | Hildebrandt et al. (2024) | 10x Visium | GSE268068 | 4 |
| Parasite | Trypanosoma brucei | Tbb | African trypanosomiasis | Brain | Quintana et al. (2022) | 10x Visium | GSE200642 | 3 |

All raw and processed data are available from Figshare: <https://doi.org/10.6084/m9.figshare.31839988>.


## Curated genes and gene sets

Table S2 summarizes 12 curated resources. Counts are shown as `Human/Mouse`.

| Type | Name | Main use | Count |
|---|---|---|---|
| Gene | Infection_Immunity | Infection and immune response genes | 391/345 |
| Gene | PRR | Pattern recognition receptor genes | 137/128 |
| Gene | Macrophage | Macrophage polarization/function genes | 35/35 |
| Gene set | GO_BP_Detect_viral | GO biological process terms related to viral infection | 9/12 |
| Gene set | GO_BP_Detect_bacterial | GO biological process terms related to bacterial infection | 10/7 |
| Gene set | KEGG_Detect_viral | KEGG viral infection pathways | 21/21 |
| Gene set | KEGG_Detect_bacterial | KEGG bacterial infection pathways | 11/7 |
| Gene set | KEGG_Detect_parasitic | KEGG parasitic infection pathways | 6/6 |
| Gene set | KEGG_Metabolism | KEGG metabolism pathways | 85/85 |
| Gene set | Reactome_Metabolism | Reactome metabolism pathways | 301/253 |
| Gene set | MSigDB_Hallmark | MSigDB hallmark gene sets | 50/50 |
| Gene set | PCD | Programmed cell death gene sets | 18/18 |

```{r gene-geneset, message = FALSE, warning = FALSE}
STID::Gene_Geneset
```

## Session information

```{r session-info}
sessionInfo()
```
