---
title: "Plotting functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  fig.retina = 2
)

set.seed(123)
```

## Overview

This vignette introduces the public plotting functions whose names start with `Plot_` in the STID analysis scripts:

- `Plot_Spatial()` for spatial visualization of discrete labels or continuous scores;
- `Plot_SpatialCoLoc()` for visualizing two-variable spatial co-localization;
- `Plot_NicheCellComm()` for visualizing niche cell-cell communication results;
- `Plot_DistLine_Exp()` for plotting expression trends along distance from ROI/niche centers;
- `Plot_DistLine_Ratio()` for plotting cell-type ratio trends along distance from ROI/niche centers.

> **Note:** The examples are shown with `eval = FALSE` because STID objects, metadata keys, niche keys, CellChat results, gene names, and output folders are dataset-specific. Replace the placeholder values before running the examples.

## Prerequisites

```{r packages, message = FALSE, warning = FALSE, eval = FALSE}
library(tidyverse)
library(Seurat)
library(STID)
```

Depending on which plotting function is used, the following packages may also be required.

```{r optional-packages, eval = FALSE}
# Common plotting/data manipulation dependencies
library(dplyr)
library(ggh4x)
library(scales)

# Cell communication visualization
library(CellChat)
library(nichenetr)
```


## Plotting-function map

| Function | Main purpose | Typical use |
|---|---|---|
| `Plot_Spatial()` | Create spatial scatter plots from an STID object or a custom data frame | Visualize cell types, spot classes, gene scores, or continuous metadata on tissue coordinates |
| `Plot_SpatialCoLoc()` | Visualize spatial co-localization of two groups, genes, or metadata features | Check whether two cell groups or two expression features overlap in niche/ROI regions |
| `Plot_NicheCellComm()` | Plot CellChat-based niche cell communication results | Inspect interaction counts, interaction weights, signaling roles, pathway bubbles, and spatial signaling maps |
| `Plot_DistLine_Exp()` | Plot gene or metadata-feature expression along distance from ROI/niche centers | Examine how expression changes from ROI center toward the edge and outside regions |
| `Plot_DistLine_Ratio()` | Plot cell-type proportions along distance from ROI/niche centers | Examine how cell-type composition changes across binned distance layers |

## Spatial plots

`Plot_Spatial()` creates spatial coordinate plots with either discrete or continuous coloring. It can read metadata directly from an STID object, or it can plot a custom data frame if coordinate columns and a grouping column are supplied.

```{r plot-spatial-stid-discrete, eval = FALSE}
# Plot default metadata from an STID object
p <- Plot_Spatial(
  STID_obj = STID_obj,
  meta_key = "coord",
  group_by = "cell_type",
  datatype = "discrete",
  pt_size = 1.1,
  title = "Cell type distribution"
)

p
```

```{r plot-spatial-stid-continuous, eval = FALSE}
# Plot a continuous score with percentile clipping
p <- Plot_Spatial(
  STID_obj = STID_obj,
  meta_key = "coord",
  group_by = "Pathogen_Score",
  datatype = "continuous",
  vmin = "p1",
  vmax = "p99",
  black_bg = TRUE,
  title = "Pathogen score"
)

p
```

```{r plot-spatial-custom-data, eval = FALSE}
# Plot from a custom data frame
p <- Plot_Spatial(
  plot_data = my_spatial_df,
  x_colnm = "x",
  y_colnm = "y",
  group_by = "ROI_label",
  facet_grpnm = "sample",
  datatype = "discrete",
  pt_size = 1.2
)

p
```

## Spatial co-localization plots

`Plot_SpatialCoLoc()` visualizes co-localization for exactly two variables. The variables can be two groups from a `group_by` column, two genes from `features`, two metadata features from `feature_colnm`, or a combination of gene and metadata features. When `niche_key` is supplied, the function focuses on niche cells; otherwise, it can use metadata through `meta_key`.

```{r plot-spatial-coloc-cell-groups, eval = FALSE}
# Co-localization of selected cell groups inside a niche
Plot_SpatialCoLoc(
  STID_obj = STID_obj,
  loop_id = "LoopAllSamp",
  niche_key = "niche_virulence",
  group_by = "cell_type",
  group_use = c("Macrophage", "Epithelial"),
  col = COLOR_LIST$PALETTE_WHITE_BG,
  pt_size = 1.5
)
```

```{r plot-spatial-coloc-features, eval = FALSE}
# Co-localization of two gene-expression features
Plot_SpatialCoLoc(
  STID_obj = STID_obj,
  loop_id = "LoopAllSamp",
  meta_key = "coord",
  features = c("GeneA", "GeneB"),
  exp_thres = 1,
  col = COLOR_LIST$PALETTE_WHITE_BG,
  pt_size = 1.5
)
```

```{r plot-spatial-coloc-metadata-features, eval = FALSE}
# Co-localization using two numeric metadata columns
Plot_SpatialCoLoc(
  STID_obj = STID_obj,
  meta_key = "M2_NicheDetect_Lasso_20240101",
  feature_colnm = c("Score_A", "Score_B"),
  exp_thres = 1,
  pt_size = 1.5
)
```

> **Note:** This function is designed for two-variable co-localization. If more than two variables are supplied, the function stops with an error.

## Niche cell communication plots

`Plot_NicheCellComm()` visualizes CellChat results generated from niche cell communication analysis. In single-sample mode, it plots interaction circle plots, heatmaps, signaling-role heatmaps, spatial signaling maps when available, and ligand-receptor bubble plots. In multi-sample mode, it compares CellChat objects across samples in a multi-sample loop.

```{r plot-niche-cellcomm-ss, eval = FALSE}
# Single-sample CellChat visualization
Plot_NicheCellComm(
  STID_obj = STID_obj,
  samp_mode = "SS",
  loop_id = "LoopAllSamp",
  CellComm_data = cellcomm_results,
  signaling = c("TGFb", "WNT"),
  sources.use = c("Epithelial"),
  targets.use = c("Immune", "Macrophage")
)
```

```{r plot-niche-cellcomm-ms, eval = FALSE}
# Multi-sample CellChat comparison
Plot_NicheCellComm(
  STID_obj = STID_obj,
  samp_mode = "MS",
  loop_id = "LoopAllMulti",
  CellComm_data = cellcomm_results,
  signaling = c("TGFb", "WNT"),
  sources.use = NULL,
  targets.use = NULL
)
```

If a custom color vector is supplied through `col`, it should be a named vector whose names match the cell-type names in the CellChat object.

```{r plot-niche-cellcomm-named-colors, eval = FALSE}
celltype_cols <- c(
  Epithelial = "#4DBBD5FF",
  Macrophage = "#E64B35FF",
  Immune = "#00A087FF"
)

Plot_NicheCellComm(
  STID_obj = STID_obj,
  samp_mode = "SS",
  CellComm_data = cellcomm_results,
  signaling = "TGFb",
  col = celltype_cols
)
```

## Expression trends along distance

`Plot_DistLine_Exp()` plots expression values along distance from ROI/niche centers. It expects a metadata table containing distance-related columns such as `ROI_edge`, `All_Dist2ROIcenter`, and `All_Dist2ROIedge`, which are typically produced by niche/ROI detection workflows.

The function can plot genes from the assay layer, numeric columns from metadata, or both.

```{r plot-distline-exp-genes, eval = FALSE}
# Plot gene-expression trends along distance from ROI center
Plot_DistLine_Exp(
  STID_obj = STID_obj,
  loop_id = "LoopAllSamp",
  meta_key = "M2_NicheDetect_Lasso_20240101",
  facet_grpnm = "sample",
  assay_id = "Spatial",
  layer_id = "data",
  features = c("GeneA", "GeneB"),
  smooth_method = "gam",
  exp_scale = TRUE,
  distance_scale = TRUE,
  linewidth = 1,
  col = COLOR_LIST$PALETTE_WHITE_BG
)
```

```{r plot-distline-exp-metadata, eval = FALSE}
# Plot numeric metadata features along distance from ROI center
Plot_DistLine_Exp(
  STID_obj = STID_obj,
  meta_key = "M2_NicheDetect_Lasso_20240101",
  facet_grpnm = "sample",
  feature_colnm = c("Pathogen_Score", "Inflammation_Score"),
  smooth_method = "loess",
  exp_scale = TRUE,
  distance_scale = TRUE
)
```

Use `distance_scale = TRUE` to scale distance by the 95th percentile within each sample. Use `exp_scale = TRUE` to standardize expression or score values within each sample and feature.

## Cell-type ratio trends along distance

`Plot_DistLine_Ratio()` summarizes spot/cell composition by distance bins and plots the fraction of selected cell types across distance layers. The bin width is controlled by `coord_interval_ratio`, which multiplies the sample-specific spatial coordinate interval.

```{r plot-distline-ratio-basic, eval = FALSE}
# Plot cell-type ratio trends across distance bins
Plot_DistLine_Ratio(
  STID_obj = STID_obj,
  loop_id = "LoopAllSamp",
  meta_key = "M2_NicheDetect_Lasso_20240101",
  group_by = "cell_type",
  facet_grpnm = "sample",
  celltypes = c("Tcell", "Bcell", "Macrophage"),
  coord_interval_ratio = 5,
  linewidth = 1,
  col = COLOR_LIST$PALETTE_WHITE_BG
)
```

```{r plot-distline-ratio-all-celltypes, eval = FALSE}
# Plot all cell types in the grouping column
Plot_DistLine_Ratio(
  STID_obj = STID_obj,
  meta_key = "M2_NicheDetect_Lasso_20240101",
  group_by = "cell_type",
  facet_grpnm = "sample",
  celltypes = NULL,
  coord_interval_ratio = 5
)
```

The shaded region and dashed line represent the median ROI-edge distance bin for each sample, helping compare composition inside and outside the ROI boundary.

## Suggested workflow

The plotting functions are often used after spot detection, ROI/niche detection, and downstream niche analysis.

```{r suggested-workflow, eval = FALSE}
# 1. Visualize spatial metadata or scores
Plot_Spatial(
  STID_obj = STID_obj,
  meta_key = "coord",
  group_by = "cell_type",
  datatype = "discrete"
)

# 2. Visualize two-variable co-localization
Plot_SpatialCoLoc(
  STID_obj = STID_obj,
  niche_key = "niche_virulence",
  group_by = "cell_type",
  group_use = c("Macrophage", "Epithelial")
)

# 3. Plot communication results from CalSampCellComm()
Plot_NicheCellComm(
  STID_obj = STID_obj,
  CellComm_data = cellcomm_results,
  signaling = c("TGFb", "WNT")
)

# 4. Plot expression changes along ROI-center distance
Plot_DistLine_Exp(
  STID_obj = STID_obj,
  meta_key = "M2_NicheDetect_Lasso_20240101",
  facet_grpnm = "sample",
  features = c("GeneA", "GeneB")
)

# 5. Plot cell-type composition changes along ROI-center distance
Plot_DistLine_Ratio(
  STID_obj = STID_obj,
  meta_key = "M2_NicheDetect_Lasso_20240101",
  group_by = "cell_type",
  facet_grpnm = "sample",
  coord_interval_ratio = 5
)
```

## Session information

```{r session-info}
sessionInfo()
```


