---
title: "Utility functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Utility functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  fig.retina = 2
)

set.seed(123)
```

## Overview

This vignette introduces four utility functions used for environment configuration, cross-format data conversion, and annotation parsing in STID workflows. `configure_conda()` prepares the Python environment required by Python-backed conversion steps. `h5ad2rds()` and `rds2h5ad()` convert data between AnnData `h5ad` files and Seurat-compatible R objects. `parse_gtf()` extracts structured gene annotation information from GTF files.

## Function summary

| Function | Purpose | Typical input | Typical output | Notes |
|---|---|---|---|---|
| `configure_conda()` | Configure the Python environment through `reticulate` | Conda environment name, or default Miniconda setup | Activated Python/Conda environment | Run before Python-backed conversion workflows |
| `h5ad2rds()` | Convert AnnData `h5ad` files to Seurat objects | `.h5ad` file path | Seurat object and/or `.rds` file | Supports spatial and single-cell modes, reductions, coordinates, and optional SCT conversion |
| `rds2h5ad()` | Convert Seurat objects to AnnData-compatible `h5ad` files | Seurat object | `.h5ad` file | Current helper supports the SeuratDisk conversion path |
| `parse_gtf()` | Parse GTF annotation files into tidy tables | `.gtf` file path | Data frame containing genomic features and attributes | Use `fil_label` to filter feature types such as `gene` |

## Prerequisites

```{r packages, message = FALSE, warning = FALSE, eval = FALSE}
library(tidyverse)
library(Seurat)
library(STID)
```

## `configure_conda()`

`configure_conda()` configures Python through `reticulate`. If no Conda environment name is supplied, it attempts to use or install the default Miniconda environment. If a Conda environment name is supplied, it checks whether that environment exists before activating it.

```{r configure-conda, eval = FALSE}
# Use default Miniconda
configure_conda()

# Use an existing Conda environment
configure_conda(conda_nm = "scanpy_env")
```

Run this configuration before using Python-backed conversion workflows such as `h5ad2rds()` and `rds2h5ad()`.

## `h5ad2rds()`

`h5ad2rds()` converts an AnnData `h5ad` file into a Seurat object. It supports spatial and single-cell modes, optional reductions, optional spatial coordinates, and optional SCT conversion.

```{r h5ad-to-rds, eval = FALSE}
stRNA <- h5ad2rds(
  file_path = "data/sample_spatial.h5ad",
  data_type = "stRNA",
  convert_mode = "scanpy",
  assay_id = "Spatial",
  X_index = "rawX",
  binsize = 100,
  SCT_index = FALSE,
  reduction_index = TRUE,
  image_index = TRUE,
  return_object = TRUE,
  grp_nm = "sample_spatial",
  dir_nm = "M0_h5ad2rds"
)
```

## `rds2h5ad()`

`rds2h5ad()` converts a Seurat object into an AnnData-compatible `h5ad` file using the SeuratDisk path.

```{r rds-to-h5ad, eval = FALSE}
rds2h5ad(
  seurat_obj = stRNA,
  data_type = "stRNA",
  convert_mode = "seurat",
  assay_id = "Spatial",
  grp_nm = "sample_spatial",
  dir_nm = "M0_rds2h5ad"
)
```

> **Note:** The Scanpy mode of `rds2h5ad()` is not implemented in the current helper. Use `convert_mode = "seurat"` for this direction.

## `parse_gtf()`

`parse_gtf()` reads a GTF file and returns a tidy data frame. The `fil_label` argument filters the feature column, for example to keep only genes.

```{r parse-gtf, eval = FALSE}
gtf_gene <- parse_gtf(
  gtf_file = "reference/annotation.gtf",
  fil_label = "gene"
)

head(gtf_gene)
```

```{r parse-gtf-use-case, eval = FALSE}
gene_table <- parse_gtf("reference/annotation.gtf", fil_label = "gene") %>%
  dplyr::select(seqname, start, end, strand, gene_id, gene_name, gene_biotype)
```

## Session information

```{r session-info}
sessionInfo()
```
